Underdispersion in the context of community athletics

### 2024-05-02 inst/parkrun_idea.Rmd
more thoughts on exponential strengths

@@

## Now 10 in-silico races:
```{r doinsilico,cache=FALSE}
alphatrue <- 2.3
betatrue <- 11 
set.seed(1)
D <- replicate(17,which(rrace(setNames(f(26,alphatrue),letters))==letters[betatrue]))
sort(D)
plot(table(D))
```

```{r trylikeinsilico,cache=FALSE}
likelihood <- apply(L[,D],1,prod) + 0.00
likelihood <- matrix(likelihood,length(alpha_try),length(beta_try))
```

```{r showlike}
round(alpha_try,2)
matplot(alpha_try,likelihood,type='b')
matplot(beta_try,t(likelihood),type='b')
matplot(alpha_try,log(likelihood)-log(max(likelihood)),type='b',ylim=c(-20,0))
matplot(beta_try,log(t(likelihood))-log(max(likelihood)),type='b',ylim=c(-20,0))
filled.contour(alpha_try,beta_try,(likelihood))
table(likelihood==0)
plot(f(26,alphatrue))
abline(v=betatrue)
contour(alpha_try,beta_try,likelihood)
abline(v=alphatrue)
abline(h=betatrue)
contour(alpha_try,beta_try,log(likelihood) - log(max(likelihood)),levels=-(0:10))
abline(v=alphatrue)
abline(h=betatrue)
```

>>>>

Ranked data arises whenever multiple entities are sorted by some
criterion; the preferred interpretation is a race in which the
sufficient statistic is the order of competitors crossing the
finishing line.  The Plackett--Luce likelihood function
[@luce1959;@plackett1975] is a generalization of the Bradley--Terry
model [@bradley1952] in which non-negative strengths are assigned to
each competitor; without loss of generality, if the finishing order is
$1,2,\ldots,n$ and competitor $i$ has strength $p_i\geqslant 0$, then
the Plackett--Luce likelihood function will be proportional to

\begin{equation}\label{plackettluce}
\prod_{i=1}^n\frac{p_i}{\sum_{j=i}^np_j}
\end{equation}

\noindent where the strengths are normalised so $\sum p_i=1$.  Here we
consider Plackett--Luce likelihood functions in which
$p_i\propto\alpha^{-i}$ for some (generally unknown) $\alpha\geqslant
1$.  The reason for the minus sign is so that, if $\alpha\gg 1$, the
rank of competitor $i$ will be close to $i$.  Sometimes it will be
convenient to parametrize in terms of $\beta=\alpha^{-1}$.

Here, observations are restricted to knowledge of the rank of a single
competitor.  Such observations are common in practice: examples would
include an educational context in which a known number of people
attempt an examination and the only information one actually has is
one's own rank ["you came fifteenth out of a class of 85"]:
confidentiality considerations mean that knowledge of classmates'
performance is censored.  One might complete a foot race and learn
that you placed, say, 194${}^\mathrm{th}$ out of 340 entries.  In this
case the identities and abilities of the other runners are of no
interest.  Other examples might include Formula 1 motor racing, in
which a driver might race in multiple seasons in fields of 22-23:
@hankin2024 notes that Sergio Perez has raced against a total of 73
distinct competitors, rendering conventional Plackett-Luce techniques
intractable.

# Simple cases

### Two competitors {-}
The simplest non-trivial case would be two competitors with strengths
proportional to $1,\alpha^{-1}$, the unit-sum constraint meaning that
the strengths are
$\left(\frac{\alpha}{1+\alpha},\frac{1}{1+\alpha}\right)$; these are
the probabilities of competitor 1 coming first and second.

### Three competitors {-}

With three competitors we will do the general case first; let's say
the strengths are $a,b,c, a+b+c=1$.

Competitor $a$ comes first means the order was either $a\succ b\succ
c$ or $a\succ c\succ b$, that is with probability

$$
\frac{a}{a+b+c}\cdot\frac{b}{b+c} + 
\frac{a}{a+b+c}\cdot\frac{c}{b+c}
=\frac{a}{a+b+c}
$$

Competitor $a$ comes second means $b\succ a\succ c$ or $c\succ a\succ b$, that is

$$
\frac{b}{a+b+c}\cdot\frac{a}{a+c} + 
\frac{c}{a+b+c}\cdot\frac{a}{a+b}
$$

Competitor $a$ comes third means $b\succ c\succ a$ or $c\succ b\succ a$, that is

$$
\frac{b}{a+b+c}\cdot\frac{c}{a+c} + 
\frac{c}{a+b+c}\cdot\frac{b}{a+b}
$$

Now the strengths of $a,b,c$ are are proportional to
$1,\alpha^{-1},\alpha^{-2}$ respectively, so they are
$\left(\frac{\alpha^2}{1+\alpha+\alpha^2},\frac{\alpha}{1+\alpha+\alpha^2},\frac{1}{1+\alpha+\alpha^2}\right)$.

**Competitor ${\mathbf a}$** comes first with probability
$\frac{\alpha^2}{1+\alpha+\alpha^2}$.  Second with probability

$$
\frac{\alpha(1+2\alpha^2+\alpha^3)}{(1+\alpha)(1+\alpha^2)(1+\alpha+\alpha^2)}
$$

[interesting observation: this has a maximum at $a\simeq 1.35271>1$ of
about $0.3466 > \frac{1}{3}$].  He comes last with a probability of

$$
\frac{1+\alpha+2\alpha^2}{(1+\alpha)(1+\alpha^2)(1+\alpha+\alpha^2)}
$$

**Competitor ${\mathbf b}$** comes first with probability
$\frac{\alpha}{1+\alpha+\alpha^2}$.  Second with probability (after
simplification) of

$$
\frac{1+\alpha^3}{(1+\alpha)(1+\alpha+\alpha^2)}
$$

and third with probability $\frac{\alpha}{1+\alpha+\alpha^2}$.

**Competitor ${\mathbf c}$** comes first with probability
$\frac{1}{1+\alpha+\alpha^2}$.  Second with probability (after
simplification) of

$$
\frac{\alpha(1+2\alpha+\alpha^3)}{(1+\alpha)(1+\alpha^2)(1+\alpha+\alpha^2)}
$$

and last with probability

$$
\frac{\alpha^3(2+\alpha+\alpha^2)}{(1+\alpha)(1+\alpha^2)(1+\alpha+\alpha^2)}
$$

@@

## Now some in-silico races

```{r doinsilico,cache=FALSE}
alphatrue <- 1.3
betatrue <- 11 
set.seed(1)
D <- replicate(7,which(rrace(setNames(f(26,alphatrue),letters))==letters[betatrue]))
sort(D)
plot(table(D))
```

```{r trylikeinsilico,cache=FALSE}
likelihood <- apply(0.00+L[,D],1,prod) + 0.00
likelihood <- matrix(likelihood,length(alpha_try),length(beta_try))
```

```{r showlike}
round(alpha_try,2)
matplot(alpha_try,likelihood*1e14,type='b')
matplot(beta_try,t(likelihood)*1e14,type='b')
matplot(alpha_try,log(likelihood)-log(max(likelihood)),type='b',ylim=c(-20,0))
matplot(beta_try,log(t(likelihood))-log(max(likelihood)),type='b',ylim=c(-20,0))
filled.contour(alpha_try,beta_try,(likelihood)*1e14)
table(likelihood==0)
plot(f(26,alphatrue))
abline(v=betatrue)
contour(alpha_try,beta_try,likelihood)
abline(v=alphatrue)
abline(h=betatrue)
contour(alpha_try,beta_try,log(likelihood) - log(max(likelihood)),levels=-(0:10))
abline(v=alphatrue)
abline(h=betatrue)
```

```{r try2}
likelihood <- apply(L[,c(11,18)],1,prod)
likelihood <- matrix(likelihood,length(alpha_try),length(beta_try))
contour(alpha_try,beta_try,likelihood)
contour(-1/alpha_try,beta_try,likelihood)
filled.contour(
    alpha_try,beta_try,log(likelihood)-max(log(likelihood )),
    color.palette = function(n) hcl.colors(n, "YlOrRd", rev = TRUE)
)

```
